Synthesizing Symmetric Lenses

### 2019-01-28 papers/SymmetricLenses/main.tex
fix nits

While domain-specific languages for writing lenses can facilitate building
principled data transformations, \textit{synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for synthesis: their specialized, sub-Turing-complete syntax
and expressive type systems drastically
reduce the space of possible programs, making search-based synthesis easier. 
%
Recent work~\cite{optician} has exploited this observation, demonstrating that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, the bijective synthesis algorithm will find a
bijective lens matching the examples, if one exists. Often,
it can find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between related data formats. 
Hence, if the synthesis algorithm can
find any bijection at all, it is likely to be the right one.

However, many real-world data formats share just part of their information
content with other representations.  For instance, two related
database tables might share four out of five columns, but each might have a
different fifth column. 
Or, one ad-hoc system configuration file might include some format-specific meta
data, such as a date or a reference number, whereas the same configuration
file on another operating system does not. 
Indeed of the 48 benchmark problems described in Section \ref{sec:evaluation},
all of the benchmarks taken from Flash Fill~\cite{flashfill} and many of the
benchmarks that synchronize between two ad hoc file formats have this
characteristic, 13 benchmarks in total. Consequently, synthesizing only strict
bijections is often not useful in practice.

@@

With simple symmetric lenses and our SRE-based likelihood measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likelihood measure.
The interesting complications arise from the need to deal with
regular-expression equivalences.  There are infinitely many regular
expressions equivalent to a given one, and the lens returned by a
type-directed search will depend which of the infinitely many possible
representations are chosen for its source and target formats.  Moreover,
there may be {\em no} lens connecting the 
source and target formats selected by the user: in general, the synthesis
algorithm has to search to find equivalent ones for which a suitable lens
exists.  
To tame this complexity, we divide the synthesis algorithm into two
communicating search procedures (Figure~\ref{fig:high-level-algorithm}),
following~\citet{optician}. 
The first, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second, \GreedySynth, takes a pair of SREs as input
and uses a greedy, type-directed algorithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The whole synthesis algorithm terminates when comparing the
likilihood score of the best lens found so far with 
the complexity of newly generated SREs 
suggests that further searching is unlikely to be fruitful.

@@

This definition differs from classic symmetric lenses in that it does not
involve a \emph{complement}.  We give a detailed comparison in
\S\ref{sec:relationship}.

@@

If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (Later, we will use undecorated variables for
stochastic regular expressions; we mark plain REs with overbars.)  When
clear from context, we shoren ``simple symmetric lens'' to just
``lens.''

@@

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and performs an identity mapping on data
matching $\BRegex$. 
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens fully connects the corresponding data in the two formats:
when one is updated, the other gets the exact same value. Both the \CreateR
and \CreateL functions are the identity function ($\CreateR \App s = s$),
and the put functions merely return the first argument
($\PutROf{s_1}{s_2} = s_2$).
%
In our example, when a \lstinline{name} is updated in one file, the identity
lens ensures it is also updated in the other file.

In contrast, the \Disconnect lens does not propagate any edits
from one format to the other.  It takes four arguments: two
regular expressions and two strings. The regular expressions specify the formats
of the corresponding data on the two sides while the strings provide default
values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ company
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateROf{\String} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). In our extended
example, the \lstinline{salary} field is only present in management files,
so we use the disconnect lens to ensure salary edits do not cause updates to the
HR file.
The defaults are strings that indicate the information is missing.

@@

Finally, there are a number of combinators that compositionally construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$ or $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, the composite lens \lstinline{name_lens} in the running
example
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
will transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the forward direction.
\bcp{We don't actually give a nontrivial example of how disconnect works.}

@@

A challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is another valid lens (where \lstinline{rep($\BRegex$)} denotes some string in the
language of $\BRegex$). Furthermore, adding the input-output example in
Figure~\ref{fig:minimized-representations} 
helps, but only a little: It may, in general, require {\em many} examples to rule
out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats. 
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

The expected number of bits required to recover data is the well-known information-theoretic concept 
of the \emph{entropy} of the data~\cite{Shannon1948}.
Calculating entropy requires a probability distribution over the
space of possible values for the data. Specifically, given a set $S$ and a
probability distribution $P \OfType S \rightarrow \Reals$ over $S$, the
entropy is $-\Sigma_{s\in  S}P(s)\Log_2P(s)$.

@@

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the highest
likelihood by performing a type-directed search. In more detail, \GreedySynth
first normalizes the stochastic regular expressions provided by \Expand into
\emph{stochastic DNF regular expressions}. These stochastic DNF regular
expressions are, intuitively, stochastic regular expressions with disjunctions
distributed over concatenations, where associativity information has been
removed. Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions and \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names). \GreedySynth
converts yields a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

@@

Our algorithm must resolve a three-way tension between the quality of
the inferred lens, the number of examples the user must provide to
eliminate unwanted lenses, and the amount of time it takes to return a
result.  For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time
searching for an even better lens.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize the likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we must first develop a probabilistic language
model for our language. We do this with \emph{stochastic regular expressions},
regular expressions annotated with probability information. With this, we can
jointly express a language and a probability distribution over that language. In
the following grammar, $\String$ ranges over strings, and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.
$$\Regex{},\RegexAlt{} \GEq{}  \String  \; | \; \emptyset \; | \; \PRegexStar{\Regex}{p} \; | \; \RegexConcat{\Regex_1}{\Regex_2} \; | \; \PRegexOr{\Regex_1}{\Regex_2}{p}$$

In the stochastic regular expression
$\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$, the annotation $\Probability$
represents the likelihood the string is in $\Regex_1$. In the stochastic regular
expression $\PRegexStar{\Regex}{\Probability}$, the annotation \Probability how
much more likely having a string involving more than $n$ iterations is than
having a string involving exactly $n$ iterations. The semantics of SREs is
defined by the function $P$ below.

@@

\subsection{Stochastic Regular Expression Equivalences}
$\Expand$ needs to be able to find equivalent stochastic regular expressions.
However, existing work does not provide a means to reason about stochastic
regular expression equivalence. We extend the
\emph{star-semiring}~\cite{Droste2009} equivalences to stochastic regular
expressions (Figure~\ref{fig:ssr-sre}).

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}
\begin{proof}
  The proofs of this theorem, and subsequent theorems in this paper, appears in
  the appendix present in the full version of the paper accompanying this
  submission.
\end{proof}

@@

\subsection{Stochastic Regular Expression Entropy}
The entropy of a data source $S$ is the expected number of bits required to
describe a random element in $S$ ($-\Sigma_{s\in S}P(s)\Log_2P(s)$). The entropy
of a stochastic regular expression can be computed directly from its syntax,
when the stochastic regular expression is unambiguous and contains no empty
subcomponents.
\begin{center}
  \begin{tabular}{rcl}
    $\EntropyOf{s}$
    & =
    & 1\\

@@

    $\EntropyOf{\PRegexOr{\Regex}{\RegexAlt}{\Probability}}$
    & =
    & $\Probability(\EntropyOf{\Regex} - \Log_2(\Probability)) + (1-\Probability)(\EntropyOf{\RegexAlt} - \Log_2(1-\Probability))$\\
  \end{tabular}
\end{center}
We have proven that this syntactic computation corresponds to the entropy of the
stochastic regular expression.

@@

Without unambiguity constraints it is not easy to calculate entropy (and the
method we have given is certainly incorrect).  For example, consider the SRE
$\PRegexOr{\lstinline{"a"}}{\lstinline{"a"}}{.5}$. Using the above computation,s
this SRE has an entropy of $1$, but the entropy should be $0$. These
difficulties become particularlly complex when the languages are ambiguous in
subtle ways, the algorithm must be correct when only some of the two languages
overlap. Similar issues occur when trying to find the entropy when $\emptyset$
is a subterm (what should the entropy of $\PRegexStar{\emptyset}{.5}$ be?),
so we do not define entropy on $\emptyset$.

@@

We now provide the typing derivations and semantics for these lenses. We do
not include the semantics for combinators equivalent to those presented in
prior work~\cite{symmetric-lenses,Boomerang}. Furthermore, we only present
\CreateR and \PutR when \CreateL and \PutL are symmetric; full details can
be found in an appendix\ifappendices\else{} (in the supplemental
materials)\fi. For simplicity of presentation, we use unambiguous regular
expressions everywhere, even though some lenses can actually be defined over
ambiguous regular expressions (for example, the identity lens doesn't need
any ambiguity constraints\bcp{Is this the only example?}); we believe
relaxing this restriction is unproblematic.

\begin{centermath}
  \begin{tabular}{m{1cm}@{\hspace*{6em}}m{1cm}@{\hspace*{3em}}}
    $
  \inferrule*
  {
  }
  {
    \IdentityLensOf{\BRegex} \OfType \BRegex \Leftrightarrow \BRegex
  }$
&
$\begin{array}[b]{r@{\ }c@{\ }l}
    \CreateR{} \App s & = & s\\
    \PutR{} \App s_1 \App s_2 & = & s_1
  \end{array}$
\end{tabular}
\end{centermath}
Note that the identity lens ignores the second argument in the put functions.
Because the two formats are fully synchronized, no knowledge of the prior data
is needed.

@@

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
$
\inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
  {
    \IterateLensOf{\Lens} \OfType
    \StarOf{\BRegex}
    \Leftrightarrow
    \StarOf{\BRegexAlt}
  }
  $
  &
  $
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
  {
    \InvertOf{\Lens} \OfType \BRegexAlt \Leftrightarrow \BRegex
  }
  $
\end{tabular}
\end{centermath}
%\begin{center}
%  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
%    $\CreateR{} \App \StringAlt$ & = & $\Lens.\CreateLOf{\StringAlt}$\\
%    $\CreateL{} \App \String$ & = & $\Lens.\CreateROf{\String}$\\
%    $\PutR{} \App \StringAlt \App \String$ & = & $\Lens.\PutLOf{\StringAlt}{\String}$\\
%    $\PutL{} \App \String \App \StringAlt$ & = & $\Lens.\PutROf{\String}{\StringAlt}$
%  \end{tabular}
%\end{center}
The \IterateLens lens deals with iterated data, while inverting reverses the direction of a lens: creating on the right becomes creating on the left and vice versa, and putting on the right becomes putting on the left and vice versa. The invert combonator is particularly useful when chaining many compositions together, as it can be used to align the central types.
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]

\subsection{Lens Costs}
Our cost metric is based on the expected amount of information required to
recover a string in one data format from the other. We use the function
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ to calculate bounds of the
expected amount of information required to recover a string in $\RegexAlt$ from
a string in $\Regex$, synchronized by $\Lens$. We use the function
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ to calculate bounds of the
expected amount of information required to recover a string in $\Regex$ from a
string in $\RegexAlt$, synchronized by $\Lens$.
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\RegexAlt \Given \IdentityLensOf{\RegexAlt}, \RegexAlt}$
    & =
    & [0,0]\\

@@

    $\REntropyOf{\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt} \Given \MergeLOf{\Lens_1}{\Lens_2}, \Regex}$
    & =
    & $(0,\REntropyOf{\RegexAlt \Given \Lens_1, \Regex_1}+\REntropyOf{\RegexAlt \Given \Lens_2, \Regex_2}+1)$\\

>>>>

While domain-specific languages for writing lenses can facilitate building
principled data transformations, \textit{synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for synthesis: their specialized, sub-Turing-complete syntax
and expressive type systems drastically
reduce the space of possible programs, making search-based synthesis easier. 
%
Recently, Optician~\cite{optician} has exploited this observation, demonstrating that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, the bijective synthesis algorithm will find a
bijective lens matching the examples, if one exists. Often,
it can find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between related data formats. 
Hence, if the synthesis algorithm can
find any bijection at all, it is likely to be the right one.

However, many real-world data formats share just part of their information
content with other representations.  For instance, two related
database tables might share four out of five columns, but each might have a
different fifth column. 
Or, one ad-hoc system configuration file might include some format-specific meta
data, such as a date or a reference number, whereas the same configuration
file on another operating system does not. 
Indeed among the 48 benchmark problems described in Section \ref{sec:evaluation},
all of the benchmarks taken from Flash Fill~\cite{flashfill} and many of the
benchmarks that synchronize between two ad hoc file formats have this
characteristic, 13 benchmarks in total. Consequently, synthesizing only strict
bijections is often not useful in practice.

@@

With simple symmetric lenses and our SRE-based likelihood measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likelihood measure.
The interesting complications arise from the need to deal with
regular-expression equivalences.  There are infinitely many regular
expressions equivalent to a given one, and the lens returned by a
type-directed search will depend on which of the infinitely many possible
representations are chosen for its source and target formats.  Moreover,
there may be {\em no} lens connecting the 
source and target formats selected by the user: in general, the synthesis
algorithm has to search to find equivalent ones for which a suitable lens
exists.  
To tame this complexity, we divide the synthesis algorithm into two
communicating search procedures (Figure~\ref{fig:high-level-algorithm}),
following~\citet{optician}. 
The first, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second, \GreedySynth, takes a pair of SREs as input
and uses a greedy, type-directed algorithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The whole synthesis algorithm terminates when comparing the
likilihood score of the best lens found so far with 
the complexity of newly generated SREs 
suggests that further searching is unlikely to be fruitful.

@@

This definition differs from classical symmetric lenses in that it does not
involve a \emph{complement}.  We give a detailed comparison in
\S\ref{sec:relationship}.

@@

If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (Later, we will use undecorated variables for
stochastic regular expressions; we mark plain REs with overbars.)  When
clear from context, we shorten ``simple symmetric lens'' to just
``lens.''

@@

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and performs an identity mapping on data
matching $\BRegex$. 
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens fully connects the corresponding data in the two formats:
when one is updated, the other gets the exact same value. Both the \CreateR
and \CreateL functions are the identity function ($\CreateR \App s = s$),
and the put functions merely return the first argument
($\PutROf{\String}{\StringAlt} = \String$).
%
In our example, when a \lstinline{name} is updated in one file, the identity
lens ensures it is also updated in the other file.

In contrast, the \Disconnect lens does not propagate any edits
from one format to the other.  It takes four arguments: two
regular expressions and two strings. The regular expressions specify the formats
of the corresponding data on the two sides while the strings provide default
values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). In our extended
example, the \lstinline{salary} field is only present in management files,
so we use the disconnect lens to ensure salary edits do not cause updates to the
HR file.
The defaults are strings that indicate the information is missing.

@@

Finally, there are a number of combinators that compositionally construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$ or $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, the composite lens \lstinline{name_lens} in the running
example
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
will transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the forward direction (converting from the left format to the right).
\bcp{We don't actually give a nontrivial example of how disconnect works.}

@@

A challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is another well-typed lens (where \lstinline{rep($\BRegex$)} denotes some string in the
language of $\BRegex$). Furthermore, adding the input-output example in
Figure~\ref{fig:minimized-representations} 
helps, but only a little: It may, in general, require {\em many} examples to rule
out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats. 
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

@@

The expected number of bits required to recover data is the well-known
information-theoretic concept of the \emph{entropy} of the
data~\cite{Shannon1948}. Calculating entropy requires a probability distribution
over the space of possible values for the data. Specifically, given a set $S$
and a probability distribution $P \OfType S \rightarrow \Reals$ over $S$, the
entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$.

@@

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the highest
likelihood by performing a type-directed search. In more detail, \GreedySynth
first normalizes the stochastic regular expressions provided by \Expand into
\emph{stochastic DNF regular expressions}. These stochastic DNF regular
expressions are, intuitively, stochastic regular expressions with disjunctions
distributed over concatenations, where associativity information has been
removed. Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions and \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names). \GreedySynth
yields a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

@@

Our algorithm must resolve a three-way tension among the quality of
the inferred lens, the number of examples the user must provide to
eliminate unwanted lenses, and the amount of time it takes to return a
result.  For example, if the algorithm has already found the lens in
Figure~\ref{fig:example_lens}, we don't want to spend a lot of time
searching for an even better lens.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize the likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we must first develop a probabilistic language
model for our language. We do this with \emph{stochastic regular expressions}~\cite{stoch-rnn,stoch-def},
regular expressions annotated with probability information. With this, we can
jointly express a language and a probability distribution over that language. In
the following grammar, $\String$ ranges over strings, and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.
$$\Regex{},\RegexAlt{} \GEq{}  \String  \; | \; \emptyset \; | \; \PRegexStar{\Regex}{p} \; | \; \RegexConcat{\Regex_1}{\Regex_2} \; | \; \PRegexOr{\Regex_1}{\Regex_2}{p}$$

In the stochastic regular expression
$\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$, the annotation $\Probability$
represents the likelihood that the string is in $\Regex_1$. In the stochastic regular
expression $\PRegexStar{\Regex}{\Probability}$, the annotation \Probability
represents how
much more likely having a string involving more than $n$ iterations is than
having a string involving exactly $n$ iterations. The semantics of SREs is
defined by the function $P$ below.

@@

\subsection{Stochastic Regular Expression Equivalences}
$\Expand$ needs to be able to find equivalent stochastic regular expressions.
However, existing work does not provide a means of reasoning about stochastic
regular expression equivalence. We extend the
\emph{star-semiring}~\cite{Droste2009} equivalences to stochastic regular
expressions (Figure~\ref{fig:ssr-sre}).

\begin{theorem}
  If $\Regex \SSREquiv \RegexAlt$ then $\ProbabilityOf{\Regex}{\String} =
  \ProbabilityOf{\RegexAlt}{\String}$, for all strings $\String \in \LanguageOf{\Regex}$.
\end{theorem}
\begin{proof}
  The proofs of this theorem, and subsequent theorems in this paper, appears in
  the appendix presented in the full version of the paper accompanying this
  submission.
\end{proof}

@@

\subsection{Stochastic Regular Expression Entropy}
The entropy of a data source $S$ is the expected number of bits required to
describe a random element in $S$ (formally $-\Sigma_{s\in S}P(s)\Log_2P(s)$). The
entropy of a stochastic regular expression can be computed directly from its
syntax, when the stochastic regular expression is unambiguous and contains no
empty subcomponents.
\begin{center}
  \begin{tabular}{rcl}
    $\EntropyOf{s}$
    & =
    & 0\\

@@

    $\EntropyOf{\PRegexOr{\Regex}{\RegexAlt}{\Probability}}$
    & =
    & $\Probability(\EntropyOf{\Regex} - \Log_2(\Probability)) + (1-\Probability)(\EntropyOf{\RegexAlt} - \Log_2(1-\Probability))$\\
  \end{tabular}
\end{center}
We have proven that this syntactic computation corresponds to the entropy of the
stochastic regular expression when the stochastic regular expression is
unambiguous (parses strings uniquely).

@@

Without unambiguity constraints it is not easy to calculate entropy (and the
method we have given is certainly incorrect).  For example, consider the SRE
$\PRegexOr{\lstinline{"a"}}{\lstinline{"a"}}{.5}$. Using the above computation,
this SRE has an entropy of $1$, but the entropy should be $0$. These
difficulties become particularlly complex when the languages are ambiguous in
subtle ways and the algorithm must be correct when only some of the two languages
overlap. Similar issues occur when trying to find the entropy when $\emptyset$
is a subterm (what should the entropy of $\PRegexStar{\emptyset}{.5}$ be?),
so we do not define entropy on $\emptyset$.

@@

We now provide the typing rules and semantics for these combinators. We do
not include the semantics for combinators equivalent to those presented in
prior work~\cite{symmetric-lenses,Boomerang}. Furthermore, we only present
\CreateR and \PutR when \CreateL and \PutL are symmetric; full details can
be found in an appendix\ifappendices\else{} (in the supplemental
materials)\fi. For simplicity of presentation, we use unambiguous regular
expressions everywhere, even though some lenses can actually be defined over
ambiguous regular expressions (for example, the identity lens doesn't need
any ambiguity constraints\bcp{Is this the only example?}); we believe
relaxing this restriction is unproblematic.

\begin{centermath}
  \begin{tabular}{m{1cm}@{\hspace*{6em}}m{1cm}@{\hspace*{3em}}}
    $
  \inferrule*
  {
  }
  {
    \IdentityLensOf{\BRegex} \OfType \BRegex \Leftrightarrow \BRegex
  }$
&
$\begin{array}[b]{r@{\ }c@{\ }l}
    \CreateR{} \App s & = & s\\
    \PutR{} \App s \App t & = & s
  \end{array}$
\end{tabular}
\end{centermath}
Note that the identity lens ignores the second argument in the put functions.
Because the two formats are fully synchronized, no knowledge of the prior data
is needed.

@@

\begin{centermath}
\begin{tabular}[b]{l@{\qquad}l}
$
\inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
  {
    \IterateLensOf{\Lens} \OfType
    \StarOf{\BRegex}
    \Leftrightarrow
    \StarOf{\BRegexAlt}
  }
  $
  &
  $
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt
  }
  {
    \InvertOf{\Lens} \OfType \BRegexAlt \Leftrightarrow \BRegex
  }
  $
\end{tabular}
\end{centermath}
%\begin{center}
%  \begin{tabular}{@{}r@{\ }c@{\ }l@{}}
%    $\CreateR{} \App \StringAlt$ & = & $\Lens.\CreateLOf{\StringAlt}$\\
%    $\CreateL{} \App \String$ & = & $\Lens.\CreateROf{\String}$\\
%    $\PutR{} \App \StringAlt \App \String$ & = & $\Lens.\PutLOf{\StringAlt}{\String}$\\
%    $\PutL{} \App \String \App \StringAlt$ & = & $\Lens.\PutROf{\String}{\StringAlt}$
%  \end{tabular}
%\end{center}
The \IterateLens lens deals with iterated data, while inverting reverses the direction of a lens: creating on the right becomes creating on the left and vice versa, and putting on the right becomes putting on the left and vice versa. The invert combinator is particularly useful when chaining many compositions together, as it can be used to align the central types.
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]

Type equivalence enables a lens of type $S \Leftrightarrow T$ to be used as a
lens of type $S' \Leftrightarrow T'$ if $S$ is equivalent to $S'$ and $T$ is
equivalent to $T'$. Type equivalence is useful both for addressing type
annotations, and for making well-typed compositions.

\subsection{Lens Costs}
Our cost metric is based on the expected amount of information required to
recover a string in one data format from the other. We use the function
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ to calculate bounds of the
expected amount of information required to recover a string in $\RegexAlt$ from
a string in $\Regex$, synchronized by $\Lens$. We use the function
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ to calculate bounds of the
expected amount of information required to recover a string in $\Regex$ from a
string in $\RegexAlt$, synchronized by $\Lens$. We use the notation $a(b,c)$ to
mean $(a*b,a*c)$, and the notation $(a,b)+(c,d)$ to mean $(a+c,b+d)$.
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\RegexAlt \Given \IdentityLensOf{\RegexAlt}, \RegexAlt}$
    & =
    & [0,0]\\

@@

    $\REntropyOf{\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt} \Given \MergeLOf{\Lens_1}{\Lens_2}, \Regex}$
    & =
    & $(0,\REntropyOf{\RegexAlt_1 \Given \Lens_1, \Regex}+\REntropyOf{\RegexAlt_2 \Given \Lens_2, \Regex}+1)$\\
